The Role of Variability in Learning Generalization: A Computational Modeling Approach

### 2024-05-28 Presentation/_fig_creation.qmd
update slides

>>>>

library(ggplot2)

# Generate distance values
distance <- seq(0, 10, length.out = 100)

# Define the c parameter values
c_values <- c(0.08, 0.5, 1)

# Calculate similarity for each c value
similarity_data <- data.frame(
  distance = rep(distance, length(c_values)),
  c = factor(rep(c_values, each = length(distance))),
  similarity = exp(-rep(c_values, each = length(distance)) * distance^2)
)

# Create the faceted plot
c_example <- ggplot(similarity_data, aes(x = distance, y = similarity)) +
  geom_line(linewidth = 1) +
  labs(x = "Distance", y = "Similarity") +
  #scale_color_manual(values = c("blue", "green", "red")) +
  facet_wrap(~ c, ncol = 3, labeller = labeller(c = function(x) paste("c =", x))) +
  theme_minimal() +
  theme(
    # plot.title = element_text(hjust = 0.5, size = 14, face = "plain"),
    # axis.title = element_text(size = 12),
    legend.position = "none",
    strip.text = element_text(size = 12)
  ) +
  ggtitle("Gaussian Similarity Decay")

ggsave(here::here("Presentation/assets/c_example.png"), c_example, width = 4.0, height = 3, units = "in", dpi = 600)

### 2024-05-28 Presentation/_hold_slides.qmd
update slides

@@

>>>>

## designs that seemingly overcome these issues

- Kerr & Booth 1978

::: footer
@kerrSpecificVariedPractice1978
:::

@@

:::: {.columns}
::: {.column width="70%"}

![](assets/sol_space.png)

![](assets/full_space.png)
:::
::: {.column width="30%"}
### Model Steps

## Computational Model - Does Similarity Account for Effect of Variability?

- Test whether similarity explains the difference in performance between Constant and Varied conditions
  - Assuming equivalent generalization (1 c for both groups) - similarity improves model fit, but does not explain group difference
  - Assume training condition influences generalization gradient (fit c separately) - now similarity does explain the difference between Constant and Varied.

::: footer
**Equation 1:** $$ Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}}) $$
**Equation 2:** $$ d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 } $$
:::

## Project 1 General Discussion {.smaller .r-fit-text}

------------------------------------------------------------------------

### 2024-05-28 Presentation/slides.qmd
update slides

-   Schema Theory (Schmidt, 1975)
-   Desirable Difficulties Framework (Bjork & Bjork, 2011)
-   Challenge Point Framework (Guadagnoli & Lee, 2004)

<!-- ![](assets/fig-toy-model1-1-1.png) -->

## Overview of Current Work {.smaller}

### Project 1[^1]

@@

::: footer
[^1]: Now published in Cognitive Psychology [@gormanInstancebasedModelAccount2022]
:::
--

# Project 1  - Issues with previous research

@@

@@

## designs that seemingly overcome these issues

- Kerr & Booth 1978

::: footer
@kerrSpecificVariedPractice1978
:::

@@

- Conceptual replication of Kerr & Booth design 
- Testing classic pattern as well

@@

@@

@@

@@

@@

## Project 1 - Discussion

-   Varied training group's superior performance
-   Computational modeling results with IGAS
-   Theoretical implications

@@

### Model Steps

- each of the 6 positions has an empirical solution space
- Compute the summed similarity between training throws, and solution spaces
  - separately for each participant, and each of the testing positions
- We now have a measure of how similar training behavior was to testing solutions

@@

:::: {.columns}
::: {.column width="70%"}

![](assets/sol_space.png)

![](assets/full_space.png)
:::
::: {.column width="30%"}
### Model Steps

- each of the 6 positions has an empirical solution space
- Compute the summed similarity between training throws, and solution spaces
  - separately for each participant, and each of the testing positions
- We now have a measure of how similar training behavior was to testing solutions

- $d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 }$
- $Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}})$
:::
::::

## Computational Model - Does Similarity Account for Effect of Variability?

- Test whether similarity explains the difference in performance between Constant and Varied conditions
  - Assuming equivalent generalization (1 c for both groups) - similarity improves model fit, but does not explain group difference
  - Assume training condition influences generalization gradient (fit c separately) - now similarity does explain the difference between Constant and Varied.

::: footer
**Equation 1:** $$ Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}}) $$
**Equation 2:** $$ d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 } $$
:::

## Project 1 General Discussion {.smaller .r-fit-text}

# Project 2 - Variability and Extrapolation in a Function Learning Task

@@

::: {.columns}
::: {.column width="50%"}
![](assets/htw_e2_test1.png){width="75%"}
![](assets/htw_e2_test3.png){width="75%"}
:::

::: {.column width="50%"}
![](assets/htw_e2_test2.png){width="75%"}
![](assets/htw_e2_test4.png){width="75%"}
:::
:::

@@

## Accuracy

@@

## Discrimination

>>>>

## Overview {.smaller}

### Project 1

@@

## Project 1 
<br>

### An instance-based model account of the benefits of varied practice in visuomotor skill[^1]

<br><br><br>

::: footer
[^1]: Now published in Cognitive Psychology -  [@gormanInstancebasedModelAccount2022]
:::

<br>

::: {.fragment .semi-fade-out}
- Desirable Difficulties Framework (Bjork & Bjork, 2011)
<br>  
- Challenge Point Framework (Guadagnoli & Lee, 2004)
::: 
- Schema Theory (Schmidt, 1975)

::: notes
Variety of theoretical explanations
Schema theory in paticular has been extremely influential 
:::

## Issues with previous research

@@

::: fragment
:::: columns
::: {.column}

![](assets/agg_issues.png)
:::
::: {.column}
<br>

-  If generalization decays exponentially with distance, constant groups 1 and 2 may have far less generalization potential to the testing condition. 
- performance of individual constant groups often not reported in analyses.

::::
:::
::: footnote
e.g. [@mccrackenTestSchemaTheory1977; @chuaPracticeVariabilityPromotes2019; @delreyEffectsContextualInterference1982]
:::

@@

## Designs that avoid similarity issue

### Kerr & Booth 1978

:::: columns
::: {.column}
![](assets/kerr_booth.png)
:::
::: {.column}
- bean bag throwing task
- Constant and varied conditions train from distinct positions
- Both groups are tested from the position where the constant group trained
- Impressive demonstration of varied training outperforming constant training 
:::
::::
::: footer
@kerrSpecificVariedPractice1978
:::

@@

- Conceptual replication of Kerr & Booth design 
- Also test positions novel to both groups
- Validate computerized visuomotor learning task

@@

::: notes
- Use mouse to flick the ball towards the target
- Must release from orange square
- Subjects have flexibility in terms of trajectory and release point inside the square
- Nicety of computerized task - we record the precise details easily
:::

@@

::: notes
Varied training significantly better at the position that was novel for all subjects. Also better even at the position that the constant group trained at.

4 different positions, the varied group trained from the easiest and the hardest position, they also interpolated the constant group. 
:::

@@

::: notes
task mostly the same - minor adjustments to barrier
6 different constant groups
varied group doesn't train on either the easiest or the hardest position

larger design to give us more data to work with for modelling
:::

@@

::: notes

- This time we can compare varied and constant training - by just comparing the varied group to
the 2 constant groups that trained at the same locations
- Here we have all 6 of the constant groups collapsed together
- also did more fine grained grained comparisons

@@

### Instance encoding, and similarity-based generalization assumptions

- instances refers to individual throws (x and y velocity)
- similarity refers to the distance between training throws and the solution space of each of the eventual testing positions.

@@

### Computing Similarity

- Euclidean distance between each training throw, and each solution space
  - separately for each participant, and each of the testing positions
- Similarity computed as a Gaussian decay function of distance, i.e. larger distances result in lower similarity
- Each participant gets their own similarity score for each of the 6 testing positions
  - similarity score for a given testing position is the sum of the similarities between each training throw and the entire empirical solution space

@@

## Non-linear similarity

![](assets/c_example.png)

## Does this similarity metric work?

- Our similarity measure does is a significant predictor of testing performance. 
  - Remains significant when controlling for training accuracy, and training condition.

## Accounting for group-level effect {.smaller}

::: fragment
- The base version of the model fit a single $c$ parameter 
  - thus assuming that everyone generalizes to the same degree
:::

::: fragment
- We next fit a 2-c version, separately optimizing for varied and constant groups, resulting in new 2-c similarity scores for each participant. 
  - The optimal c value for the varied group is smaller than that of the constant group, indicative of broader generalization. 
:::

::: fragment
- When 2-c similarity is added to the linear model predicting testing performance as a function of condition, the effect of training condition is no longer significant. 
  - not the case with the original 1-c similarities. 
:::

::: fragment  
- Thus the influence of varied training can be explained by an instance-based similarity model, IF one assumes flexibility in the generalization gradient. 
:::

## Conclusions

::: fragment 
- New empirical support for a benefit of variability in visuomotor skill tasks
:::

::: fragment 
- Some support for the less common pattern observed by @kerrSpecificVariedPractice1978
:::

::: fragment 
- Demonstrated the utility of a similarity based approach in this domain
:::

# Project 2

## Project 2

### Impact of Training Variability on Visuomotor Function Learning and Extrapolation

@@

::: {.columns}
::: {.column width="50%"}
![](assets/htw_e2_test1.png){width="75%" .fragment fragment-index=1}
![](assets/htw_e2_test3.png){width="75%" .fragment fragment-index=2}
:::

::: {.column width="50%"}
![](assets/htw_e2_test2.png){width="75%" .fragment fragment-index=1}
![](assets/htw_e2_test4.png){width="75%" .fragment fragment-index=2}
:::
:::

@@

## Experiment 3 - Accuracy

@@

## Experiment 3 - Discrimination
